---
title: "Explanatory Data Analysis"
author: "Sofia Rivas and Jocelyn Rojas"
format: pdf
editor: visual
---

# 1. Load Relevant Packages

```{r}
#| output: false

library(tidyverse) #data wrangling (includes other packages)
library(dplyr) #data wrangling 
library(ggplot2) #plotting
library(brms) #Bayesian statistics 
library(bayesplot) #posterior probability distributions 
library(factoextra) #variance plot (scree plot)
library(tidybayes) #helps convert Bayesian modeling for ggplot2 and tidyverse
library(patchwork) #put plots together
```

# 2. Data Wrangling

explain where the data came from and why we select certain columns

```{r}
#raw data
predictors_raw <- readRDS("quad_build3.rds") 

#select relevant columns and delete rows with NA values 
predictors <- predictors_raw %>% 
   mutate(juveniles = macj + nerj + ptej + lsetj + eisj,
         n_macro_plants_m2 = n_macro_plants_20m2/20,
         macro_stipe_density_m2 = macro_stipe_density_20m2/20,
         adults = (density20m2_ptecal + density20m2_eisarb + density20m2_nerlue + density20m2_lamset)/20,
         marine_snails = tegula_densitym2 + pomaulax_densitym2,
         density_cancer = density20m2_cancer_spp/20,
         density_lamstump = density20m2_lamstump/20,
         density_macstump = density20m2_macstump/20) %>% 
  dplyr::select(mean_gi, n_macro_plants_m2, macro_stipe_density_m2, adults,
juveniles, density_cancer, density_lamstump, density_macstump, relief_cm, risk_index, purple_urchin_densitym2, purple_urchin_conceiledm2, red_urchin_densitym2, red_urchin_conceiledm2, marine_snails, lamr, macr, cov_crustose_coralline, cov_desmarestia_spp, cov_lam_holdfast_live, cov_articulated_coralline, cov_fleshy_red, cov_bare_rock, cov_green_algae, cov_barnacle, cov_bare_sand, cov_colonial_tunicate, cov_dictyota_dictyopteris, cov_sponge, cov_dead_kelp_holdfast_any, cov_mac_holdfast_live, cov_red_turf_2_cm, cov_corynactis_californica, cov_dictyoneurum_spp) %>% 
  na.omit()

predictors2 <- predictors %>%
  dplyr::select(mean_gi, cov_mac_holdfast_live, macro_stipe_density_m2, n_macro_plants_m2, cov_bare_sand, purple_urchin_conceiledm2, cov_crustose_coralline, purple_urchin_densitym2, relief_cm, marine_snails, red_urchin_densitym2, cov_corynactis_californica, cov_barnacle, cov_articulated_coralline, cov_desmarestia_spp, cov_colonial_tunicate, density_lamstump, cov_dead_kelp_holdfast_any, cov_dictyoneurum_spp, red_urchin_conceiledm2)
  
```

# 3. Principal Component Analysis

-   Explain why we chose PCA

## Step 1: PCA

```{r}
#run PCA: run only on predictor variables, not the response variable (mean_gi)
#standardize variables 
pca <- prcomp(predictors %>% 
                dplyr::select(-mean_gi), 
                scale. = TRUE) 

#visualize 
fviz_eig(pca)
```

```{r}
#create matrix of raw PC scores (chosen from scree plot)
pc_scores <- as.data.frame(pca$x[, 1:10])

#add response variable (mean_gi) into new dataframe
pc_data <- pc_scores %>%
              mutate(mean_gi = predictors$mean_gi)

#set priors for brm 
priors_pca <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(6.5, 2), class = "Intercept"), #mean mean_gi is ~6.5
  prior(student_t(3, 0, 2), class = "sigma"))

#run brm model
model_pca <- brm(mean_gi ~ .,
                 data = pc_data,
                 family = gaussian(),
                 prior = priors_pca,
                 chains = 4, iter = 4000, warmup = 2000,
                 cores = 4)
summary(model_pca) #PC1, PC3, and PC6 are significant!
```

explain why we switch to skew_normal family (include pp_checks and loos)

```{r}
#posterior predictive check 
pp_check(model_pca) #mostly fits well, doesn't account for bump around x=2

#final model 
model_pca_skew <- brm(mean_gi ~ PC1 + PC3 + PC6,  # only significant PCs
                   data = pc_data,
                   family = skew_normal(), #changed to skew_normal
                   prior = priors_pca,
                   chains = 4, iter = 4000, warmup = 2000,
                   cores = 4)
summary(model_pca_skew) #again, PC1, PC3, and PC6 are significant! 
bayes_R2(model_pca_skew) #check R^2

#quick check to make sure skew_normal() is a better fit than gaussian()
pp_check(model_pca_skew)

loo_gaussian <- loo(model_pca)
loo_skew <- loo(model_pca_skew, reloo = TRUE)
loo_compare(loo_gaussian, loo_skew)

#what predictors are included in PC1, PC3, and PC6?
for(pc in c("PC1","PC3","PC6")) { #starts a for loop 
  cat("\n---", pc, "---\n") #syntax for output: headers for each PC
  cat("TOP POSITIVE:\n") #syntax for output: label for the variables with  largest positive loadings for the current PC
  print(head(sort(pca$rotation[,pc], decreasing = TRUE), 5)) #sorts loadings from largest to smallest, keeps top 5 variables 
  cat("TOP NEGATIVE:\n") #syntax for output: label for the variables with  largest negative loadings for the current PC
  print(head(sort(pca$rotation[,pc], decreasing = FALSE), 5))} #sorts loadings from largest to smallest, keeps top 5 variables 
```

interpret graph (posterior distributions)

```{r}
#density curves of each PC showing posterior distributions 
mcmc_areas(model_pca_skew,
           pars = vars(starts_with("b_"), -b_Intercept), #plot regression coefficients and exclude intercept 
           prob = 0.95, #shade the 95% credible interval 
           prob_outer = 1.0) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + #dotted line at x=0 for visual assessment 
  theme_classic()
```

explain predicted vs observation plot

```{r}
pc_data$fitted <- fitted(model_pca_skew)[, "Estimate"] #posterior mean prediction for each observation
pc_data$lower  <- fitted(model_pca_skew)[, "Q2.5"] #lower 95% credible interval
pc_data$upper  <- fitted(model_pca_skew)[, "Q97.5"] #upper 95% credible interval

#plot observed vs predicted to see how well the three ecologically meaningful PCs predict gonad index
ggplot(pc_data, aes(x = fitted/100, y = mean_gi/100)) + #turn units to %
  geom_point(color = "#466995") + #each dot is one observation 
  geom_errorbarh(aes(xmin = lower, xmax = upper), alpha = 0.2, size = 0.3) + #horizontal error bars showing prediction uncertainty
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") + #prediction line 
  labs(x = "Predicted Mean GI (%)", y = "Observed Mean GI (%)") +
  scale_x_continuous(limits = c(0,0.15)) +
  scale_y_continuous(limits = c(0,0.2)) +
  theme_classic()
```

```{r}
#plotting 
fviz_pca_var(
  pca,
  col.var = "contrib",  # color by contribution to the PC axes
  repel = TRUE,
  gradient.cols = c("grey80", "steelblue", "darkblue")
) +
  labs(title = "PCA variable plot")


fviz_pca_var(pca, axes = c(1, 3),
             col.var = "contrib",
             repel = TRUE,
             gradient.cols = c("grey80", "steelblue", "darkblue")) +
  labs(title = "PCA variable plot - PC1 vs PC3")

# PC1 vs PC6
fviz_pca_var(pca, axes = c(1, 6),
             col.var = "contrib",
             repel = TRUE,
             gradient.cols = c("grey80", "steelblue", "darkblue")) +
  labs(title = "PCA variable plot - PC1 vs PC6")

# PC3 vs PC6
fviz_pca_var(pca, axes = c(3, 6),
             col.var = "contrib",
             repel = TRUE,
             gradient.cols = c("grey80", "steelblue", "darkblue")) +
  labs(title = "PCA variable plot - PC3 vs PC6")


#add individual plots for each variable  
```

# 3. Principal Component Analysis (the real one) 

-   Explain why we chose PCA

## Step 1: PCA

```{r}
#run PCA: run only on predictor variables, not the response variable (mean_gi)
#standardize variables 
pca <- prcomp(predictors2 %>% 
                dplyr::select(-mean_gi), 
                scale. = TRUE) 

#visualize 
fviz_eig(pca)
```

```{r}
#create matrix of raw PC scores (chosen from scree plot)
pc_scores <- as.data.frame(pca$x[, 1:10])

#add response variable (mean_gi) into new dataframe
pc_data <- pc_scores %>%
              mutate(mean_gi = predictors2$mean_gi)

#set priors for brm 
priors_pca <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(6.5, 2), class = "Intercept"), #mean mean_gi is ~6.5
  prior(student_t(3, 0, 2), class = "sigma"))

#run brm model
model_pca <- brm(mean_gi ~ .,
                 data = pc_data,
                 family = gaussian(),
                 prior = priors_pca,
                 chains = 4, iter = 4000, warmup = 2000,
                 cores = 4)
summary(model_pca) #PC1, PC3, and PC6 are significant!
```

explain why we switch to skew_normal family (include pp_checks and loos)

```{r}
#posterior predictive check 
pp_check(model_pca) #mostly fits well, doesn't account for bump around x=2
model_pca_gaus <- brm(mean_gi ~ PC1 + PC2 + PC3,
                                  data = pc_data,
                                  family = gaussian(),
                                  prior = priors_pca,
                                  chains = 4, iter = 4000, warmup = 2000,
                                  cores = 4)
pp_check(model_pca_gaus)

#final model 
model_pca_skew <- brm(mean_gi ~ PC1 + PC2 + PC3,  # only significant PCs
                   data = pc_data,
                   family = skew_normal(), #changed to skew_normal
                   prior = priors_pca,
                   chains = 4, iter = 4000, warmup = 2000,
                   cores = 4)
summary(model_pca_skew) #again, PC1, PC2, and PC3 are significant! 
bayes_R2(model_pca_skew) #check R^2

#quick check to make sure skew_normal() is a better fit than gaussian()
pp_check(model_pca_skew)

loo_gaussian <- loo(model_pca_gaus, moment_match = TRUE)
loo_skew <- loo(model_pca_skew, reloo = TRUE)
loo_compare(loo_gaussian, loo_skew)


model_pca_gaus$formula       # check gaussian model formula
model_pca_skew$formula  # check skew_normal model formula

#what predictors are included in PC1, PC3, and PC6?
for(pc in c("PC1","PC2","PC3")) { #starts a for loop 
  cat("\n---", pc, "---\n") #syntax for output: headers for each PC
  cat("TOP POSITIVE:\n") #syntax for output: label for the variables with  largest positive loadings for the current PC
  print(head(sort(pca$rotation[,pc], decreasing = TRUE), 5)) #sorts loadings from largest to smallest, keeps top 5 variables 
  cat("TOP NEGATIVE:\n") #syntax for output: label for the variables with  largest negative loadings for the current PC
  print(head(sort(pca$rotation[,pc], decreasing = FALSE), 5))} #sorts loadings from largest to smallest, keeps top 5 variables 
```

interpret graph (posterior distributions)

```{r}
#density curves of each PC showing posterior distributions 
mcmc_areas(model_pca_skew,
           pars = vars(starts_with("b_"), -b_Intercept), #plot regression coefficients and exclude intercept 
           prob = 0.95, #shade the 95% credible interval 
           prob_outer = 1.0) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + #dotted line at x=0 for visual assessment 
  theme_classic()

```

explain predicted vs observation plot

```{r}
pc_data$fitted <- fitted(model_pca_skew)[, "Estimate"] #posterior mean prediction for each observation
pc_data$lower  <- fitted(model_pca_skew)[, "Q2.5"] #lower 95% credible interval
pc_data$upper  <- fitted(model_pca_skew)[, "Q97.5"] #upper 95% credible interval

#plot observed vs predicted to see how well the three ecologically meaningful PCs predict gonad index
ggplot(pc_data, aes(x = fitted/100, y = mean_gi/100)) + #turn units to %
  geom_point(color = "#466995") + #each dot is one observation 
  geom_errorbarh(aes(xmin = lower/100, xmax = upper/100), 
                 alpha = 0.2, 
                 linewidth = 0.3) + #horizontal error bars showing prediction uncertainty
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") + #prediction line 
  labs(x = "Predicted Mean GI (%)", y = "Observed Mean GI (%)") +
  scale_x_continuous(limits = c(0,0.15)) +
  scale_y_continuous(limits = c(0,0.2)) +
  theme_classic()
```

```{r}

summary(pca)$importance[2, 1:3] * 100 # % variance for each PC

#plotting PC1 and PC2
fviz_pca_var(pca, axes = c(1, 2),#specify which PCs to plot
             select.var = list(contrib = 8),  #only top # contributors
             col.var = "contrib",
             repel = TRUE,
               labelsize = 2,
             gradient.cols = c("grey80", "steelblue", "darkblue")) +
  labs(title = "PCA variable plot: PC1 and PC2",
       x = "PC1 (22.7%)",
       y = "PC2 (11.9%)",
       color = "Contribution (%)")+
  theme_classic()

#add individual plots for each PC  
# Plot for PC1
p1 <- pc_data %>%
  ggplot(aes(x = PC1, y = mean_gi)) +
  stat_smooth(method = "lm", color = "steelblue", fill = "lightblue") +
  geom_point(color = "#466995", alpha = 0.6) +
  labs(x = "PC1 (kelp forest structure)", 
       y = "Mean GI (%)") +
  theme_classic()

# Plot for PC2
p2 <- pc_data %>%
  ggplot(aes(x = PC2, y = mean_gi)) +
  stat_smooth(method = "lm", color = "steelblue", fill = "lightblue") +
  geom_point(color = "#466995", alpha = 0.6) +
  labs(x = "PC2 (urchin barren vs coralline habitat)", 
       y = "Mean GI (%)") +
  theme_classic()

# Plot for PC3
p3 <- pc_data %>%
  ggplot(aes(x = PC3, y = mean_gi)) +
  stat_smooth(method = "lm", color = "steelblue", fill = "lightblue") +
  geom_point(color = "#466995", alpha = 0.6) +
  labs(x = "PC3 (algal diversity vs urchin aggregation)", 
       y = "Mean GI (%)") +
  theme_classic()

# Combine all three
p1 + p2 + p3

```

# 4. uhhhh - horseshoe?

```{r}
#explain why hand-picking is not as honest as PCA? 
glm <- glm(mean_gi ~ PC1 + PC3 + PC6, data = pc_data)


#HORSESHOE 

#scale predictors 
predictors_scaled <- predictors %>%
  mutate(across(-mean_gi, scale))

#model with horseshoe instead of pca 
brm_hs <- brm(
  mean_gi ~ .,
  data = predictors_scaled,
  prior = prior(horseshoe(1), class = "b"),
  control = list(adapt_delta = 0.99))

summary(brm_hs)

#select only significant variables
significant_vars <- c("b_adults", 
                      "b_purple_urchin_densitym2",
                      "b_purple_urchin_conceiledm2") 
mcmc_areas(brm_hs,
           pars = significant_vars,
           prob = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_classic()




fixef(brm_hs)

pp_check(brm_hs, pars = significant_vars)

conditional_effects(brm_hs)
bayes_R2(brm_hs, pars = significant_vars)
library(bayesplot)
posterior <- as_draws_df(brm_hs)
coef_cols <- grep("^b_", colnames(posterior), value = TRUE)
coef_cols <- coef_cols[coef_cols != "b_Intercept"]  # remove intercept
mcmc_intervals(posterior, pars = coef_cols)
coef_summary <- posterior %>%
  summarise(across(all_of(coef_cols), ~ median(.))) %>%
  pivot_longer(everything(), names_to = "predictor", values_to = "median")
top_predictors <- coef_summary %>%
  filter(abs(median) > 0.05) %>%  # threshold can be adjusted
  pull(predictor) %>%
  gsub("^b_", "", .)  # remove b_ prefix to match PCA loadings
loadings <- as.data.frame(pca$rotation)
top_loadings <- loadings[top_predictors, 1:4]  # first 4 PCs
top_loadings$predictor <- rownames(top_loadings)
top_loadings
long_loadings <- top_loadings %>%
  pivot_longer(cols = starts_with("PC"), names_to = "PC", values_to = "loading")




```
