---
title: "Exploratory Data Analysis"
subtitle: Environmental Drivers of Purple Urchin Gonad Index in Central California Kelp Forests
format: pdf
editor: visual
authors: Sofia Rivas & Jocelyn Rojas
---

## Environmental Drivers of Purple Urchin Gonad Condition

### Project Overview

#### Research Question

To what extent do biological and environmental variables predict sea urchin gonad index across kelp forests and urchin barrens, and which variables are the strongest predictors of gonad index?

#### Background

Project background can be found in our "Explanatory Data Analysis" files.

#### Target Audience:

Recovery and restoration practitioners and researchers

## Exploratory Data Analyses

```{r}
# Loading Relevant Packages
library(MASS)
library(performance)
library(tidyverse) #data wrangling (includes other packages)
library(dplyr) #data wrangling 
library(ggplot2) #plotting
library(brms) #Bayesian statistics 
library(bayesplot) #posterior probability distributions 
library(factoextra) #variance plot (scree plot)
library(tidybayes) #helps convert Bayesian modeling for ggplot2 and tidyverse
library(patchwork) #put plots together
```

#### Data Wrangling

```{r}
gonadindex_raw <- readRDS("quad_build3.rds")
## write.csv(gonadindex_raw, "gonadindex_raw.csv", row.names = FALSE)

## Cleaning the data ----

# New data table is gi_predictors
gi_predictors <- gonadindex_raw %>%
  # Grouping site details into one variable called "site_id"
  mutate(site_id = paste(site, pred_patch, zone, year, sep = " ")) %>%
  # Removing listed variables from the tibble
  select(-c(patch_id, latitude, longitude, survey_date, patch_cat, 
            zone, year, sd_biomass, sd_gi, se_gonad_mass_g, geometry)) %>% 
  # Grouping all juvenile kelp data into one variable called "juveniles"
  mutate(juveniles = macj + nerj + ptej + lsetj + eisj,
         adults = density20m2_ptecal + density20m2_eisarb + density20m2_nerlue) %>% 
  # Changing area data to 1 m2
  mutate (n_macro_plants_m2 = n_macro_plants_20m2/20)
```

#### GLM Model Testing

We received guidance from Sofia's academic advisor, Joshua Smith, to determine the type of model that would be most appropriate for answering our research question. He suggested we run a Generalized Linear Model (glm). Our data frame contained over 70 predictor variables so we brainstormed variable combinations when writing out our models.

After running each model, we used the summary() and r2() functions to determine the AIC and r2 values. We were advised to play close attention to each model's r squared value.

```{r}
# We named our first model m1 and included the following predictor variables: total_biomass_g, purple_urchin_densitym2 + juveniles

m1 <- glm(mean_gi ~ mean_gonad_mass_g +
            total_biomass_g +
            purple_urchin_densitym2 + 
            juveniles,
          data = gi_predictors, family = gaussian)

# Generating Model Summary and R^2
summary(m1) # AIC: 384.43
r2(m1) # 0.599

# Second model (m2) with the following predictor variables: mean_gonad_mass_g, total_biomass_g, purple_urchin_densitym2, juveniles, macr, cov_crustose_coralline, cov_mac_holdfast_live, lamr, n_macro_plants_m2

m2 <- glm(mean_gi ~ mean_gonad_mass_g +
            total_biomass_g +
            purple_urchin_densitym2 + 
            juveniles + 
            macr + 
            cov_crustose_coralline + 
            cov_mac_holdfast_live +
            lamr +
            n_macro_plants_m2 
          , data = gi_predictors, family = gaussian)

# Generating Model Summary and R^2
summary(m2) # AIC: 375.79
r2(m2) # 0.674

# Third model (m3) with the following predictor variables:mean_gonad_mass_g, total_biomass_g, purple_urchin_densitym2, juveniles, lamr, n_macro_plants_m2

m3 <- glm(mean_gi ~ mean_gonad_mass_g + 
            total_biomass_g + 
            purple_urchin_densitym2 + 
            juveniles + 
            lamr + 
            n_macro_plants_m2
          , data = gi_predictors, family = gaussian)

## After running the first and second models, we realized that our mean_gi variable included "mean_gonad_mass_g" and "total_biomass_g" in its calculation so we decided to remove them from the rest of our exploratory models

# Generating Model Summary and R^2
summary(m3) # AIC: 371.21
r2(m3) # 0.668

# Fourth model (m4) with the following predictor variables: purple_urchin_densitym2 + #ecological drivers only juveniles, lamr, n_macro_plants_m2, macr, cov_mac_holdfast_live,  cov_crustose_coralline

m4 <- glm(mean_gi ~ purple_urchin_densitym2 + 
            juveniles +
            lamr +
            n_macro_plants_m2 +
            macr +
            cov_mac_holdfast_live +
            cov_crustose_coralline,
          , data = gi_predictors, family = gaussian)

# Generating Model Summary and R^2
summary(m4) # AIC: 421.07
r2(m4) # 0.436

# Fifth model (m5) with the following predictor variables: purple_urchin_densitym2, lamr, n_macro_plants_m2, macr, adults

m5 <- glm(mean_gi ~ purple_urchin_densitym2 +
            lamr +
            n_macro_plants_m2 +
            macr +
            adults 
          , family = gaussian, data = gi_predictors)

# Generating Model Summary and R^2
summary(m5) # AIC: 413.67
r2(m5) # 0.457

# We used the StepAIC function to test variations of each of our exploratory models 

# AIC Model 1
AIC.m1 <- stepAIC(m1,
                  direction = "both", 
                  trace = TRUE)
# AIC Model 2
AIC.m2 <- stepAIC(m2,
                  direction = "both", 
                  trace = TRUE)
# AIC Model 3
AIC.m3 <- stepAIC(m3,
                  direction = "both", 
                  trace = TRUE)
# AIC Model 4
AIC.m4 <- stepAIC(m4,
                  direction = "both", 
                  trace = TRUE)
# AIC Model 5
AIC.m5 <- stepAIC(m5,
                  direction = "both", 
                  trace = TRUE)
```

We continued to run similar models with different combinations of predictors but the five that are included above had the highest r2 values. However, the highest r2 of all our exploratory models was only 0.674. We determined a glm may not be the best model for our data and decided to start over by plotting the distribution of our variable of interest: mean_gi

#### Distribution Testing

We wanted to check if our data was followed a Gaussian distribution so we ran the following lines of code.

```{r}
# Plotting the distribution of "mean_gi" data
hist(gi_predictors$mean_gi)
```

```{r}
# Produces a Q-Q plot to assess whether the data follows a normal distribution
# If data (more or less) follows a straight line, it is normally distributed
qqnorm(gi_predictors$mean_gi)

# Adds a line to Q-Q plot
qqline(gi_predictors$mean_gi)
```

Our data was normally distributed with a slight right skew, so we decided to run a Bayesian alternative to a Linear Mixed Effects Model using the **brms** package. Our data was collected from different sites and within each site, sampling took place across various site types.

Site (Location) –\> Site Type (Kelp forest, Urchin Barren, and Intermediate Zone) –\> Site Depth (Shallow/Deep)

#### Bayesian Model Testing

```{r}
# Our first exploratory Bayseian model included the density of two different species of kelp (major purple urchin food source), site type, and site location as a random effect

library(brms)

# BRMS model
mean_gi_model <- brm (data = gi_predictors,
             family = gaussian(),
             mean_gi ~ 0 + lamr + macr + site_type + (1 | site),
             iter = 2000, warmup = 1000, chains = 4, cores = 4, 
             seed = 4, 
             file = "mean_gi_model")

# We used the summary() and plot() function to check: 
  # Rhat values, confidence intervals, and chains. 
```

Our Rhat values were 1, our confidence intervals did not include zero, the plots all centered around a single point, and our chains overlapped. However, our data frame included over 70 predictor variables. We realized we needed a way to determine what variables were the best to include in our final model.

After consulting with other students, we learned about Principal Component Analysis, a statistical method used to identify the most predictive variables in a complex data set.

#### Principal Component Analysis Testing

```{r}
# To run a PCA, we needed to make all of our values numeric
# We created a new data frame called gi_numeric to keep the other values as they were

gi_numeric <- gi_predictors %>%
  dplyr::select(where(is.numeric)) %>%
  na.omit()

# PCA ATTEMPT 1 
pca <- prcomp(gi_numeric %>% dplyr::select(-mean_gi), 
              scale. = TRUE) 

# To determine the number of PCAs we needed to run, we used the "factoextra" package and fviz_eig function to plot a scree plot. The scree plot is a line graph used in PCA to determine how many factors to retain.

library(factoextra)
fviz_eig(pca) # Function to plot a scree plot

# Based on the scree plot, we needed to run 7 PCAs
pc_scores <- as.data.frame(pca$x[, 1:7]) 

# Added the response variable (mean_gi) back in
pc_data <- pc_scores %>%
  mutate(mean_gi = gi_numeric$mean_gi)

```

#### Bayesian Model Testing 2

```{r}
# After running our PCA, we rewrote our BRMS model so it included our PC scores

# Priors
priors <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(6.5, 2), class = "Intercept"),
  prior(student_t(3, 0, 2), class = "sigma"))

#BRMS Model
model_prior <- brm(mean_gi ~ .,
                   data = pc_data,
                   family = gaussian(),
                   prior = priors,
                   sample_prior = "only",
                   chains = 4, iter = 2000)

# Evaluates the model by comparing the observed data with the simulated data 
pp_check(model_prior)  # should look like plausible mean_gi values
```

## Conclusion

After running various models to answer our research question, we determined running a BRMS model including the results from our PCA was the best approach. We also wanted to update our model with better informed priors.

Our final model and explanatory data analysis can be found in our "Explanatory Data Analysis" quarto document.
